---
title: "Compute Dynamic Stability for the Maizuru Bay Fish Community"
author: "Hao Ye"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compute Dynamic Stability for the Maizuru Bay Fish Community}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
bibliography: my_lib.bib
---

```{r, include = FALSE}
LOCAL <- identical(Sys.getenv("LOCAL"), "true")
knitr::opts_chunk$set(purl = LOCAL, 
                      collapse = TRUE,
                      comment = "#>", 
                      fig.width = 6, 
                      eval = FALSE)
```

# Introduction

This document demonstrates the dynamic stability analysis, using the Maizuru Bay fish community as an example dataset [@Ushio_2018].

First, we load the packages that we are going to use, and set the seed (for the random number generation associated with creating surrogate data):

```{r packages, message = FALSE}
library(portalDS)
library(dplyr)
library(ggplot2)

set.seed(42)
```

# The Data

First, we load in the data and take a look at the time series for the fish populations.

```{r time series}
data(maizuru_block)

maizuru_palette <- viridis::viridis(NCOL(maizuru_block) - 1)
names(maizuru_palette) <- sort(names(maizuru_block[, -1]))

x_axis_time <- scale_x_date(limits = as.Date(c("2005-03-01", "2011-01-01")),
                            date_breaks = "1 year", date_labels = "%Y", expand = c(0.01, 0))


plot_time_series(maizuru_block) +
  x_axis_time +
  scale_color_manual(values = maizuru_palette) + 
  guides(color = FALSE)
```

# Dynamic Stability Calculations

Perform the dynamic stability analysis for the maizuru system. The steps involved in the full analysis are:

1a. Run simplex projection on each time series to identify the optimal embedding dimension
1b. Generate surrogate time series (here, using the twin-surrogate method)

```{r compute simplex}
simplex_results <- compute_simplex(maizuru_block,
                                   E_list = 1:24,
                                   surrogate_method = "twin", 
                                   num_surr = 200, 
                                   surr_params = list(T_period = 24, 
                                                      quantile_vec = c(0.875, 0.88, 0.89, 0.90, 0.91, 0.92, 
                                                                       0.93, 0.94, 0.95, 0.85, 0.84, 0.83, 
                                                                       0.82, 0.81, 0.80, 0.96))
)
```

2. Run ccm on each pairwise interaction, including the surrogate data, and using the embedding dimension identified previously. Since this takes a while to run, we've commented out the actual code, and instead load a pre-generated results file.

```{r compute ccm}
# future::plan(future.callr::callr)
# ccm_results <- compute_ccm(simplex_results,
#                            lib_sizes = c(6, 12, 24, 40, 80, 140, 220, 320, NROW(maizuru_block)), 
#                            num_samples = 200)

data_path <- system.file("extdata", "maizuru_ccm_results.RDS",
                         package = "portalDS", mustWork = TRUE)
ccm_results <- readRDS(data_path)
```
    
3. identify the significant interactions by comparing the CCM results for the real time series against the comparable calculations for the surrogate data

```{r ccm links}
ccm_links <- compute_ccm_links(ccm_results)
```

4. run S-map models for each time series, using the appropriate number of lags, and including the important interacting variables (we rescale the time series to each have mean = 0, variance = 1)

```{r smap coeffs}
smap_coeffs <- compute_smap_coeffs(maizuru_block, ccm_links, 
                                   rescale = TRUE)
```

5. extract out the s-map coefficients from the models and assemble matrices for the system

```{r smap matrices}
smap_matrices <- compute_smap_matrices(smap_coeffs,
                                       ccm_links)

# add date labels for each matrix in list
stopifnot(length(smap_matrices) == NROW(maizuru_block))
names(smap_matrices) <- maizuru_block$censusdate
```

6. perform eigen-decomposition on the s-map coefficient matrices

```{r eigen-decomposition}
eigen_decomp <- compute_eigen_decomp(smap_matrices)
eigenvalues <- eigen_decomp$values
eigenvectors <- eigen_decomp$vectors
```

# Figures

## Interaction Network

First, we look at the interaction network. `plot_network` has a default layout and palette, but we can specify both. Here, we re-use the viridis color scheme used to plot the time series.

```{r interaction network}
maizuru_network <- plot_network(ccm_links, palette = maizuru_palette)
```

Note that the output of `plot_network` is a list. The `plot` element is the ggraph object, which we can plot:

```{r view interaction network}
print(maizuru_network$plot)
```

The remaining components allow us to plot a subset of the network, while keeping the same layout of nodes and the associated colors.

```{r interaction subnetwork}
halichoeres_links <- ccm_links %>%
  filter(stringr::str_detect(lib_column, "Halichoeres") | 
           stringr::str_detect(target_column, "Halichoeres"))

halichoeres_network <- plot_network(halichoeres_links, 
                                    palette = maizuru_palette, 
                                    existing_graph = maizuru_network$graph)

print(halichoeres_network$plot)
```

## Eigenvalues

Dynamic stability [@Ushio_2018] is defined as the time-varying property that is the largest (absolute value) eigenvalue. With the results of our analysis, we can plot this as a time series, applying the same x-axis transformation as for the population time series earlier:

```{r}
eigenvalues %>%
  plot_eigenvalues() +
  x_axis_time
```

Note that `plot_eigenvalues` has a number of additional parameters, which allow us to plot multiple eigenvalues simultaneously, or identify when the dominant eigenvalue is complex (red points):

```{r}
eigenvalues %>%
  plot_eigenvalues(num_values = 2, highlight_complex = TRUE) + 
  x_axis_time
```

## Eigenvectors

Just as with the eigenvalues, we can plot the dominant eigenvector(s) as it changes in time. Note that there are only 13 dimensions here, because we drop the 2 species that share no links with the others.

```{r}
eigenvectors %>%
  plot_eigenvectors(num_values = 2) + 
  scale_color_manual(values = maizuru_palette) + 
  x_axis_time
```

# References