---
title: "Rocky Intertidal Dynamic Stability Analysis"
author: "Hao Ye"
date: "`r Sys.Date()`"
output: github_document
bibliography: ../vignettes/my_lib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report documents applying the dynamic stability analysis to the rocky intertidal system reported on in [@Beninca_2015]

First, some setup for the environment:

```{r packages, message = FALSE}
library(portalDS)
library(dplyr)
library(ggplot2)

set.seed(42)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Workflow

## Data

We use the interpolated data, as these methods also require regular sampling in time.

```{r, collapse = TRUE}
block <- readxl::read_excel(here::here("inst/extdata/pnas.1421968112.sd01.xlsx"), 
                            sheet = 2, 
                            col_types = c("text", "numeric", "numeric", "numeric", "numeric", "guess", "guess", "guess")) %>%
  mutate(Date = as.Date(as.numeric(Date) - 366, origin = "0000-12-31")) %>%
  select(1:5)
names(block) <- c("date", "barnacles", "crustose_algae", "mussels", "bare_rock")
```

## Analysis

We do not go through the full analysis here. Instead, see the [Maizuru Dynamic Stability vignette](https://ha0ye.github.io/portalDS/articles/maizuru-dynamic-stability.html) or (eventual methdos write-up).

```{r}
results_file <- here::here("output/rocky_intertidal_ds_results.RDS")
results <- compute_dynamic_stability(block, results_file, id_var = "date")
str(results, max.level = 1)
```

# Results

## Abundance time series

```{r, fig.width = 10}
plot_time_series(results$block, time_column = "date")
```

## Eigenvalues & Eigenvectors

```{r}
plot_eigenvalues(results$eigenvalues, 
                 num_values = 3)
```

```{r, fig.width = 10}
plot_eigenvectors(results$eigenvectors)
```

## Singular values and SVD vectors

```{r}
plot_svd_values(results$svd_decomp$d, 
                num_values = 3)
```

```{r, fig.width = 10}
plot_svd_vectors(results$svd_decomp$u)
```

## Volume contraction and total variance

```{r}
plot_volume_contraction(results$volume_contraction)
```

```{r}
plot_total_variance(results$total_variance)
```

# References

