---
title: "Portal Dynamic Stability Analysis"
author: "Hao Ye"
date: "`r Sys.Date()`"
output: github_document
bibliography: ../vignettes/my_lib.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This report documents applying the dynamic stability analysis to the Portal dataset [@Ernest_2019].

First, some setup for the environment:

```{r packages, message = FALSE}
library(portalDS)
library(dplyr)
library(ggplot2)

set.seed(42)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Workflow

## Data

We use data from the 8 control plots, up to the 2015 treatment-switch. Abundances are summed, and scaled according to the number of plots sampled in each census (sometimes, incomplete sampling occurred). Only species with at least 50% non-zero abundance across the time series were kept. Missing censuses had abundances imputed using linear interpolation for each individual species.

```{r, collapse = TRUE}
block <- make_portal_block(filter_q = 0.5)
str(block)
```

## Analysis

We do not go through the full analysis here. Instead, see the [Maizuru Dynamic Stability vignette](https://ha0ye.github.io/portalDS/articles/maizuru-dynamic-stability.html) or (eventual methdos write-up).

We specify a results file for storing the outputs of the analysis. The `compute_dynamic_stability()` function will check for the presence of individual output components and will skip them if they've already been computed.

```{r}
results_file <- here::here("output/portal_ds_results_50.RDS")
results <- compute_dynamic_stability(block, results_file, id_var = "censusdate")
str(results, max.level = 1)
```

# Results

## Abundance time series

```{r, fig.width = 10}
plot_time_series(results$block)
```

## Eigenvalues & Eigenvectors

Highlighted segments are the posterior estimates of regime shifts from Christensen et al. 2018.

```{r}
plot_eigenvalues(results$eigenvalues, 
                 num_values = 3) %>% 
    add_regime_shift_highlight()
```

```{r, fig.width = 10}
plot_eigenvectors(results$eigenvectors) %>% 
    add_regime_shift_highlight()
```

## Singular values and SVD vectors

Again, highlighted segments are the posterior estimates of regime shifts from Christensen et al. 2018.

```{r}
plot_svd_values(results$svd_decomp$d, 
                num_values = 3) %>% 
    add_regime_shift_highlight()
```

```{r, fig.width = 10}
plot_svd_vectors(results$svd_decomp$u) %>% 
    add_regime_shift_highlight()
```

## Volume contraction and total variance

```{r}
plot_volume_contraction(results$volume_contraction) %>%
  add_regime_shift_highlight()
```

```{r}
plot_total_variance(results$total_variance) %>%
  add_regime_shift_highlight()
```

# References
